Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant 098051.}

### 2016-07-22 paper/paper.tex
Updates to introduction and conclusions

The \emph{variation graph toolkit vg} \cite{Garrison2014-2016} is a community effort to develop such tools. This paper describes \emph{GCSA2}, the path index used in vg. A \emph{path index} is a generalization of text indexes for labeled graphs. Given a query string, it finds the paths with a label matching the query. Indexing graphs for path queries is inherently hard, as the number of paths increases exponentially with path length. The design of a path index is hence a trade-off between maximum query length, the size of the index, and pruning complex regions of the graph.

@@

The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index for representing \emph{de~Bruijn graphs} space-efficiently. Rødland \cite{Roedland2013} proposed another similar generalization. Succinct de~Bruijn graphs can represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly. Compacted \cite{Cazaux2014} and compressed \cite{Marcus2014} de~Bruijn graphs achieve similar space usage by representing unary paths in the graph as single nodes. Probabilistic de~Bruijn graphs \cite{Pell2012} use Bloom filters to store the graph. They support faster queries in similar space at the expense of producing false positives. Read aligners based on various encodings of de~Bruijn graphs have been proposed \cite{Limasset2015,Liu2016}.

Some path indexes store the graph as a collection of sequences. BWBBLE \cite{Huang2013} uses the powerset alphabet for encoding simple variants and creates new sequences with a sufficient amount of context for complex variants. Queries are slower than in ordinary FM\nobreakdash-indexes because of the powerset alphabet. vBWT \cite{Maciuca2016} encodes variant sites explicitly in the sequence as  $X(A|B|C)Y$, using distinct separator symbols for each site. As the variants must be taken into account during search, queries are slower than in ordinary FM\nobreakdash-indexes. The \emph{hypertext index} \cite{Thachuk2013} assumes that the nodes of the graph are labeled with strings. The strings are indexed using an ordinary FM-index. Partial matches in the strings are combined into full matches with range queries in the edge matrix. While matches crossing one edge are easy to find, the approach becomes impractical with matches crossing multiple edges.

@@

The above methods can be classified in three categories based on the data models they use. Most of them deal with graphs arising from \emph{aligned sequences}, assuming a shared global sequence with local variation. GCSA-based methods can index directed acyclic graphs, as well as cyclic graphs that are sufficiently similar to \emph{de~Bruijn graphs}. GCSA2, the hypertext index, and indexes based on de~Bruijn graphs work with \emph{arbitrary graphs}.

GCSA2 combines ideas from the original GCSA and from succinct de~Bruijn graphs. Conceptually it uses a de~Bruijn graph as a path index of a variation graph. The \orderk{k} de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely. The pruned graph is encoded with a generalization of the FM\nobreakdash-index. GCSA2 often uses less than 1~bit per \kmer{k} with large values of $k$. Other de~Bruijn graph-based path indexes cannot achieve similar space usage, as they store some information for each \kmer{k} explicitly. GCSA2 also includes extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph. The extensions are used for e.g.~finding maximal exact matches in the vg read aligner.

@@

We built vg graphs from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}. To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial edges with \texttt{vg mod -p -l -e 4}, and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}. We extracted all paths of length $16$ from the forward strands of the graphs. There were a total of 4.80~billion paths with 1.53~billion distinct labels. We built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes.

Table~\ref{table:construction} summarizes index construction, while Table~\ref{table:indexes} contains details on the size of the indexes. We can build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache. The indexes contain roughly $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the pruned de~Bruijn graphs use only 4.4\nobreakdash--5.8~billion nodes to represent them. Index size grows even more slowly with $k$, as we sample more values in lower order graphs. For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph.

@@

GCSA2 is a path index for variation graphs. It uses a pruned de~Bruijn graph as an index of the variation graph, and encodes that with a generalization of the FM\nobreakdash-index. The index includes extensions based on suffix trees; other extensions have also been considered but not implemented yet (see Appendix~\ref{appendix:hypertext}). GCSA2 has been integrated in the variation graph toolkit vg, where it is used for e.g.~read alignment based on maximal exact matches.

@@

I thank Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions we had while I developed the GCSA2 index, and for the new ideas I got from the discussions.

>>>>

The \emph{variation graph toolkit vg} \cite{Garrison2014-2016} is a community effort to develop such tools. This paper describes \emph{GCSA2}, the path index used in vg. A \emph{path index} is a generalization of text indexes for labeled graphs. Given a query string, it finds the paths with a label matching the query. Indexing graphs is inherently hard, as the number of paths increases exponentially with path length. The design of a path index is hence a trade-off between maximum query length, index size and performance, and pruning complex regions of the graph.

@@

\emph{De~Bruijn graphs} can be used as \kmer{k} indexes of other graphs. There are already some read aligners based on them \cite{Limasset2015,Liu2016}. The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} encodes the graph space-efficiently with generalization of the FM\nobreakdash-index. Rødland \cite{Roedland2013} proposed another similar generalization. Succinct de~Bruijn graphs can simulate \orderk{k} de~Bruijn graphs for multiple values of $k$ \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly. Other space-efficient representations include compacted \cite{Cazaux2014} and compressed \cite{Marcus2014} de~Bruijn graphs, which represent unary paths in the graph as single nodes or edges. Probabilistic de~Bruijn graphs \cite{Pell2012} use Bloom filters to support faster queries, at the expense of producing false positives. All these space-efficient representations require several bits per \kmer{k} for the graph itself, and more for mapping back to the indexed graph.

Some path indexes store the graph as a collection of sequences. BWBBLE \cite{Huang2013} uses the powerset alphabet for encoding substitutions and creates new sequences with a sufficient amount of context for other variants. vBWT \cite{Maciuca2016} encodes variant sites explicitly in the sequence as $X(A|B|C)Y$, using distinct separator symbols for each site. Queries in both BWBBLE and vBWT are slower than in ordinary FM\nobreakdash-indexes, as variant sites force the search to branch. The \emph{hypertext index} \cite{Thachuk2013} works with graphs that have string labels in nodes. The strings are indexed using an FM-index. Partial matches in the strings are combined into full matches with range queries in the edge matrix. While matches crossing one edge are easy to find, the approach becomes impractical with matches crossing multiple edges.

@@

The above methods can be classified in three categories based on the data models they use. Many deal with graphs arising from \emph{aligned sequences}, assuming a shared global sequence with local variation. GCSA-based methods can index directed acyclic graphs, as well as cyclic graphs that are sufficiently similar to \emph{de~Bruijn graphs}. GCSA2, the hypertext index, and indexes based on de~Bruijn graphs work with \emph{arbitrary graphs}.

GCSA2 combines ideas from the original GCSA and from succinct de~Bruijn graphs. Conceptually it uses a de~Bruijn graph as a path index of a variation graph. The \orderk{k} de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely. The pruned graph is encoded with a generalization of the FM\nobreakdash-index. GCSA2 often uses around 1~bit per \kmer{k} with large values of $k$. Explicit de~Bruijn graphs cannot achieve similar space usage, as they store some information for each \kmer{k}. GCSA2 also includes extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph. The extensions are used for e.g.~finding maximal exact matches in the vg read aligner.

@@

We built vg graphs from the human reference genome (GRCh37) and 1000~Genomes Project variation \cite{1000GP2015}. To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial edges with \texttt{vg mod -p -l -e 4}, and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}. We extracted all paths of length $16$ from the forward strands of the graphs. There were a total of 4.80~billion paths with 1.53~billion distinct labels. We then built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes.

Table~\ref{table:construction} summarizes index construction, while Table~\ref{table:indexes} contains details on the size of the indexes. We can build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache. The indexes contain roughly $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the pruned de~Bruijn graphs use only 4.4\nobreakdash--5.8~billion nodes to represent them. Index size grows even more slowly with $k$, as we sample more values in lower order graphs. For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph. Extensions from Section~\ref{sect:extensions} increase index size to 1.12~bits per \kmer{k}.

@@

GCSA2 is a path index for variation graphs. It uses a de~Bruijn graph as an index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index. The index supports queries of length up to $k$ exactly and longer queries with false positives. GCSA2 includes extensions based on suffix trees; other extensions have also been considered but not implemented yet (see Appendix~\ref{appendix:hypertext}). The index has been integrated in the variation graph toolkit vg, where it is used for e.g.~read alignment based on maximal exact matches.

@@

The author thanks Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions while developing the GCSA2 index.
